---
title: "LPME Simulation Studies"
author: "Robert Zielinski"
date: "December 10, 2025"
format: html
---

This notebook includes code to run simulations comparing performance between the Principal Manifold Estimation (PME), Longitudinal Principal Manifold Estimation (LPME), and Principal Curve/Surface (PC/PS) algorithms on data generated from a common underlying manifold, but with varying degrees of noise added independently between time points. The LPME algorithm is designed to obtain *smooth* estimates of the manifolds underlying noisy data. Consequently, if the algorithm achieves its stated goal, it will not show the best goodness of fit to the observed data in these conditions. Simulated data is useful in this setting because we can generate the data from a known underlying manifold, allowing us to measure the algorithm's success at recovering this known manifold.

```{r, echo=FALSE}
#| label: chunk_setup
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r, echo=FALSE}
#| label: load_packages
library(tidyverse)
library(plotly)
library(pracma)
library(furrr)
library(here)
library(progress)
library(pme)
library(princurve)
library(doSNOW)
library(foreach)
library(doRNG)

setwd(here("code"))

source("prinSurf_v3.R")
source("functions/simulations/simulate_data.R")
source("functions/simulations/preprocess_data.R")
source("functions/simulations/fit_models.R")
source("functions/simulations/evaluate_models.R")
source("functions/simulations/display_results.R")
source("functions/calculate_lpme_reconstructions.R")
source("functions/calculate_pme_reconstructions.R")
```
## Simulated Manifolds

Manifold learning algorithms typically have tradeoffs when compared to each other, meaning that an algorithm that displays strong performance for a certain class of manifolds may be outperformed by another algorithm when considering manifolds that exhibit different behavior. Because of this, it is advantageous to evaluate manifold learning algorithms with several manifolds meant to test an algorithm's performance on data with a variety of attributes. The manifolds we consider are given in Table 1 of the corresponding manuscript and are generated using the `simulate_data` function found in `code/functions/simulations/simulate_data.R`.

In the function above, we consider eight different manifolds:

1. For $r \in [-3, 3]$, $f(r) = (r, \sin(r + \frac{\pi}{2}))$.
```{r, echo=FALSE}
#| label: plot_case1
set.seed(100)
sim_data <- simulate_data(
  duration = 2,
  interval = 0.5,
  case = 1,
  obs_noise = 0.05,
  amplitude_noise = 0.1,
  period_noise = 0.05,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~time,
  z = ~X2,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~time,
    z = ~X_true2,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```

2. For $r \in [-3\pi, 3\pi]$, $f(r) = (r, \sin(r))$.
```{r, echo=FALSE}
#| label: plot_case2
set.seed(200)
sim_data <- simulate_data(
  duration = 2,
  interval = 0.5,
  case = 2,
  obs_noise = 0.1,
  amplitude_noise = 0.25,
  period_noise = 0.05,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~time,
  z = ~X2,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~time,
    z = ~X_true2,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```


3. For $r \in [-\frac{4}{5}\pi, \frac{\pi}{2}]$, $f(r) = (\cos(r), \sin(r))$.
```{r, echo=FALSE}
#| label: plot_case3

set.seed(300)
sim_data <- simulate_data(
  duration = 2,
  interval = 0.5,
  case = 3,
  obs_noise = 0.15,
  amplitude_noise = 0.25,
  period_noise = 0.10,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~time,
  z = ~X2,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~time,
    z = ~X_true2,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```

4. For $r \in [-1, 1]$, $f(r) = (r, r^2, r^3)$
```{r, echo=FALSE}
#| label: plot_case4
set.seed(400)
sim_data <- simulate_data(
  duration = 2,
  interval = 0.5,
  case = 4,
  obs_noise = 0.15,
  amplitude_noise = 0.25,
  period_noise = 0.10,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~X2,
  z = ~X3,
  frame = ~time,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~X_true2,
    z = ~X_true3,
    frame = ~time,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```

5. For $r \in [0, 3\pi]$, $f(r) = (r, \cos(r), \sin(r))$

```{r, echo=FALSE}
#| label: plot_case5
set.seed(500)
sim_data <- simulate_data(
  duration = 2,
  interval = 0.5,
  case = 5,
  obs_noise = 0.15,
  amplitude_noise = 0.25,
  period_noise = 0.10,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~X2,
  z = ~X3,
  frame = ~time,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~X_true2,
    z = ~X_true3,
    frame = ~time,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```

6. For $r_1 \in [-1, 1]$, $r_2 \in [-1, 1]$, $f(r_1, r_2) = (r_1, r_2, \|\mathbf{r}\|^2)$

```{r, echo=FALSE}
#| label: plot_case6
set.seed(600)
sim_data <- simulate_data(
  duration = 2,
  interval = 0.5,
  case = 6,
  obs_noise = 0.15,
  amplitude_noise = 0.25,
  period_noise = 0.10,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~X2,
  z = ~X3,
  frame = ~time,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~X_true2,
    z = ~X_true3,
    frame = ~time,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```

7. For $r_1 \in [0, 3\pi]$, $r_2 \in [-1, 1]$, $f(r_1, r_2) = (r_1\cos(r_1), r_1\sin(r_1), r_2)$

```{r, echo=FALSE}
#| label: plot_case7
set.seed(700)
sim_data <- simulate_data(
  duration = 2,
  interval = 0.5,
  case = 7,
  obs_noise = 0.15,
  amplitude_noise = 0.05,
  period_noise = 0.01,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~X2,
  z = ~X3,
  frame = ~time,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~X_true2,
    z = ~X_true3,
    frame = ~time,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```

8. For $r_1 \in [0, \pi]$, $r_2 \in [0, 2\pi]$, $f(r_1, r_2) = (\sin(r_1)\cos(r_2), \sin(r_1)\sin(r_2), \cos(r_1))$

```{r, echo=FALSE}
#| label: plot_case8
set.seed(800)
sim_data <- simulate_data(
  duration = 2,
  interval = 0.5,
  case = 8,
  obs_noise = 0.15,
  amplitude_noise = 0.25,
  period_noise = 0.05,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

plot_ly(
  data = sim_df,
  x = ~X1,
  y = ~X2,
  z = ~X3,
  frame = ~time,
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, color = "black"),
  name = "Data"
) |>
  add_markers(
    x = ~X_true1,
    y = ~X_true2,
    z = ~X_true3,
    frame = ~time,
    marker = list(size = 3, color = "red"),
    name = "True Manifold"
  )
```

Note that case 8, in which observations drawn from a 2-dimensional manifold take the form of a sphere in 3 dimensions, is of particular interest due to our application to the surfaces of subcortical structures in the brain.

Simulation runs were conducted by varying the following arguments:

- `duration`: The simulated study's follow-up duration. Values tested are 1, 2, and 5, representing study durations of 1, 2, and 5 years, respectively.
- `interval`: The scheduled interval of study visits. Values tested are 0.25 and 0.5, representing study visits every 3 months and every 6 months, respectively.
- `case`: Specifies the manifold used to generate the data. 
- `obs_noise`: The standard deviation of the independent Gaussian observation-level noise. Values tested include {0.05, 0.1, 0.25, 0.5}.
- `amplitude_noise`: The standard deviation of Gaussian noise applied to the amplitude of the manifold at each time point. Values tested include {0, 0.05, 0.1, 0.25, 0.5}.
- `period_noise`: The standard deviation of Gaussian noise applied to the period of the manifold at each time point. Values tested include {0, 0.05, 0.1, 0.25, 0.5}.
- `time_trend`: Specifies systematic longitudinal changes in the manifold amplitude. Possible values include "constant", indicating that the manifold does not change over time, "linear", representing a constant rate of change in manifold amplitude, and "quadratic", indicating a quadratic rate of change in the manifold amplitude. To approximate trends observed when the brain atrophies, all changes in amplitude are downward.
- `time_change`: Scales the longitudinal amplitude changes. Values tested include {0, 0.05, 0.1, 0.25, 0.5}.

The sample size at each time point is set to 1,000. A factorial design was used to specify combinations of parameter values, with each simulation case being run a total of 9,000 times.

## Simulation Analysis Pipeline

For our purposes, analysis of the simulated data consists of the following steps:

- Preprocessing: Standardize observations to fall between -1 and 1 and augment with polar or spherical coordinates if needed.
- Manifold learning: Fit the LPME, PME and Principal Curve/Surface algorithms to the observed data, and measure the time required to run each method.
- Model evaluation: For each model, obtain reconstructed estimates of the original observations by projecting them to the estimated manifold. Compute the mean squared distance between these estimates and the D-dimensional observations generated from the true manifold.

Functions specifying each step of the analysis process can be found at `code/functions/simulations/preprocess_data.R`, `code/functions/simulations/fit_models.R`, and `code/functions/simulations/evaluate_models.R`. 

An additional function used to generate plots to display the resulting model estimates can be found at `code/functions/simulations/display_results.R`.

Demonstrations of this analysis pipeline for several simulation cases can be found below.

### Simulation Demonstration

```{r, eval=TRUE}
#| label: demo_case1
set.seed(900)

sim_data <- simulate_data(
  duration = 2,
  interval = 0.5,
  case = 1,
  obs_noise = 0.1,
  amplitude_noise = 0.25,
  period_noise = 0.1,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

sim_processed <- preprocess_data(sim_df, case = 1, d = 1, D = 2)
sim_models <- fit_models(sim_processed, case = 1, d = 1, D = 2)
```

```{r}
model_errors <- evaluate_models(
  sim_processed,
  sim_models,
  case = 1,
  d = 1,
  D = 2
)
model_errors
```

The model evaluation output above displays the MSD values associated with model estimates, along with the time required to fit each model. Note that the error outputs for partitioned models are missing, as this manifold is not self-intersecting and does not require partitioning. We see that the LPME estimate shows slightly improved ability to recover the true manifold in this setting when compared to the PME model applied naively at individual time points.

The plot of the manifold estimates shown below indicates that while the PME estimates reflect the time point-specific noise present in the observed data (shown in black), the LPME estimate is not as responsive to this noise. However, because the amplitude of the observed curve is highest at the latest time points, LPME does estimate an upward trend in the amplitude across all time points, causing bias in the estimates at earlier time points.

```{r}
result_plot <- display_results(
  sim_processed,
  sim_models,
  case = 1,
  d = 1,
  D = 2
)
result_plot
```
Another example of the simulation analysis workflow is shown below for spherical case 8.

```{r, eval=TRUE}
#| label: demo_case8
set.seed(1200)
sim_data <- simulate_data(
  duration = 2,
  interval = 0.5,
  case = 8,
  obs_noise = 0.1,
  amplitude_noise = 0.25,
  period_noise = 0.05,
  time_trend = "constant",
  time_change = 0,
  N = 1000
)
sim_df <- sim_data$df

sim_processed <- preprocess_data(sim_df, case = 8, d = 2, D = 3)
sim_models <- fit_models(sim_processed, case = 8, d = 2, D = 3)
model_errors <- evaluate_models(
  sim_processed,
  sim_models,
  case = 8,
  d = 2,
  D = 3
)
model_errors

result_plot <- display_results(
  sim_processed,
  sim_models,
  case = 8,
  d = 2,
  D = 3
)
result_plot
```

## Running Simulations

The code chunks below specify the combinations of simulation parameters and runs the simulations in parallel. Note that due to computational demands, these simulations are not run in the process of compiling this notebook. In practice, this was run on a high performance computing cluster. `run_simulations.sh` is used to initiate a batch job in slurm, which runs `code/run_simulations.R` in parallel. The code of `code/run_simulations.R` is shown in the chunk below. Broadly, this code creates a matrix with all permutations of the simulation parameters described above, then loops through each simulation case, running the workflow given in the examples above. The simulated data and the resulting manifold estimates are saved individually for each simulation run, to be summarized in `code/02_lpme_simulation_analysis.qmd`.

```{r, eval=FALSE}
#| label: run_simulations

# set possible values of simulation parameters as discussed above
durations <- c(1, 2, 5)
intervals <- c(0.25, 0.5)
cases <- 8:1
obs_noises <- c(0.05, 0.1, 0.25, 0.5)
amplitude_noises <- c(0, 0.05, 0.1, 0.25, 0.5)
period_noises <- c(0, 0.05, 0.1, 0.25, 0.5)
time_trends <- c("constant", "linear", "quadratic")
time_changes <- c(0, 0.05, 0.1, 0.25, 0.5)

# create dataframe of all combinations of parameter values
parameter_df <- expand_grid(
  duration = durations,
  interval = intervals,
  case = cases,
  obs_noise = obs_noises,
  amplitude_noise = amplitude_noises,
  period_noise = period_noises,
  time_trend = time_trends,
  time_change = time_changes
)

# get already-run simulations
file_names <- foreach(
  row_idx = seq_len(nrow(parameter_df)),
  .combine = c
) %do%
  {
    duration_value <- parameter_df$duration[row_idx]
    interval_value <- parameter_df$interval[row_idx]
    case_value <- parameter_df$case[row_idx]
    obs_noise_value <- parameter_df$obs_noise[row_idx]
    amplitude_noise_value <- parameter_df$amplitude_noise[row_idx]
    period_noise_value <- parameter_df$period_noise[row_idx]
    trend_value <- parameter_df$time_trend[row_idx]
    change_value <- parameter_df$time_change[row_idx]

    # additional stability is needed for Swiss roll and spherical cases
    # to maintain structural integrity
    if (case_value == 3) {
      period_noise_value <- period_noise_value / 5
    } else if (case_value == 5) {
      period_noise_value <- period_noise_value / 5
      obs_noise_value <- obs_noise_value / 5
    } else if (case_value == 7) {
      amplitude_noise_value <- amplitude_noise_value / 10
      period_noise_value <- period_noise_value / 10
    } else if (case_value == 8) {
      period_noise_value <- period_noise_value / 10
    }

    result_dir <- paste0("../output/simulations/case", case_value)
    if (!dir.exists(result_dir)) {
      dir.create(result_dir, recursive = TRUE)
    }

    simulation_file <- paste0(
      "duration_",
      duration_value,
      "_interval_",
      str_pad(as.character(interval_value * 100), 3, pad = "0"),
      "_obs_noise_",
      str_pad(as.character(obs_noise_value * 100), 3, pad = "0"),
      "_amplitude_noise_",
      str_pad(as.character(amplitude_noise_value * 100), 3, pad = "0"),
      "_period_noise_",
      str_pad(as.character(period_noise_value * 100), 3, pad = "0"),
      "_time_trend_",
      trend_value,
      "_time_change_",
      str_pad(as.character(change_value * 100), 3, pad = "0"),
      ".RDS"
    )

    file.path(result_dir, simulation_file)
  }


# do not repeat simulations that have already been run
included <- !file.exists(file_names)
parameter_df <- parameter_df[included, ]

seed_states <- list()

cl <- makeCluster(parallel::detectCores())
registerDoSNOW(cl)

registerDoRNG(611492)

pb <- txtProgressBar(max = nrow(parameter_df), style = 3)
progress <- function(n) setTxtProgressBar(pb, n)
opts <- list(progress = progress)
error_list <- foreach(
  row_idx = seq_len(nrow(parameter_df)),
  .export = c(
    "simulate_data",
    "preprocess_data",
    "fit_models",
    "evaluate_models",
    "display_results"
  ),
  .packages = c("tidyverse", "pme"),
  # .combine = rbind,
  .inorder = TRUE,
  .options.snow = opts,
  .errorhandling = "stop"
) %do%
  {
    print(paste("Running case", row_idx, "of", nrow(parameter_df)))

    seed_val <- .Random.seed
    seed_states[[row_idx]] <- seed_val

    duration_value <- parameter_df$duration[row_idx]
    interval_value <- parameter_df$interval[row_idx]
    case_value <- parameter_df$case[row_idx]
    obs_noise_value <- parameter_df$obs_noise[row_idx]
    amplitude_noise_value <- parameter_df$amplitude_noise[row_idx]
    period_noise_value <- parameter_df$period_noise[row_idx]
    trend_value <- parameter_df$time_trend[row_idx]
    change_value <- parameter_df$time_change[row_idx]

    # additional stability is needed for Swiss roll and spherical cases
    # to maintain structural integrity
    if (case_value == 3) {
      period_noise_value <- period_noise_value / 5
    } else if (case_value == 5) {
      period_noise_value <- period_noise_value / 5
      obs_noise_value <- obs_noise_value / 5
    } else if (case_value == 7) {
      amplitude_noise_value <- amplitude_noise_value / 10
      period_noise_value <- period_noise_value / 10
    } else if (case_value == 8) {
      period_noise_value <- period_noise_value / 10
    }

    D_value <- case_when(
      case_value == 1 ~ 2,
      case_value == 2 ~ 2,
      case_value == 3 ~ 2,
      case_value == 4 ~ 3,
      case_value == 5 ~ 3,
      case_value == 6 ~ 3,
      case_value == 7 ~ 3,
      case_value == 8 ~ 3
    )
    d_value <- case_when(
      case_value == 1 ~ 1,
      case_value == 2 ~ 1,
      case_value == 3 ~ 1,
      case_value == 4 ~ 1,
      case_value == 5 ~ 1,
      case_value == 6 ~ 2,
      case_value == 7 ~ 2,
      case_value == 8 ~ 2
    )

    result_dir <- paste0("../output/simulations/case", case_value)
    if (!dir.exists(result_dir)) {
      dir.create(result_dir, recursive = TRUE)
    }

    simulation_file <- paste0(
      "duration_",
      duration_value,
      "_interval_",
      str_pad(as.character(interval_value * 100), 3, pad = "0"),
      "_obs_noise_",
      str_pad(as.character(obs_noise_value * 100), 3, pad = "0"),
      "_amplitude_noise_",
      str_pad(as.character(amplitude_noise_value * 100), 3, pad = "0"),
      "_period_noise_",
      str_pad(as.character(period_noise_value * 100), 3, pad = "0"),
      "_time_trend_",
      trend_value,
      "_time_change_",
      str_pad(as.character(change_value * 100), 3, pad = "0"),
      ".RDS"
    )

    sim_data <- simulate_data(
      duration = duration_value,
      interval = interval_value,
      case = case_value,
      obs_noise = obs_noise_value,
      amplitude_noise = amplitude_noise_value,
      period_noise = period_noise_value,
      time_trend = trend_value,
      time_change = change_value,
      N = 1000
    )

    print(paste("Data generated for case", row_idx, "of", nrow(parameter_df)))

    sim_df <- sim_data$df

    sim_processed <- preprocess_data(
      sim_df,
      case = case_value,
      d = d_value,
      D = D_value
    )

    print(paste("Fitting models for case", row_idx, "of", nrow(parameter_df)))

    sim_models <- fit_models(
      sim_processed,
      case = case_value,
      d = d_value,
      D = D_value
    )

    print(paste(
      "Evaluating models for case",
      row_idx,
      "of",
      nrow(parameter_df)
    ))

    model_errors <- evaluate_models(
      sim_processed,
      sim_models,
      case = case_value,
      d = d_value,
      D = D_value
    )

    print(paste(
      "Plotting results for case",
      row_idx,
      "of",
      nrow(parameter_df)
    ))

    result_plot <- display_results(
      sim_processed,
      sim_models,
      case = case_value,
      d = d_value,
      D = D_value
    )

    print(paste(
      "Saving results for case",
      row_idx,
      "of",
      nrow(parameter_df)
    ))

    simulation_output <- list(
      data = sim_data,
      processed_data = sim_processed,
      models = sim_models,
      errors = model_errors,
      plot = result_plot,
      seed = seed_val
    )

    # save results for each simulation run separately
    # these will be aggregated in the following notebook
    saveRDS(simulation_output, file.path(result_dir, simulation_file))

    unlist(model_errors)
  }

stopCluster(cl)
```
