---
title: "ADNI MRI Preprocessing"
author: "Robert Zielinski"
date: "December 10, 2025"
format: html
---

This notebook describes the MRI preprocessing steps taken before applying LPME to estimate subcortical surfaces. The code below is included in `code/adni_mri_preprocess.R`, which was run as a batch job by `adni_processing.sh`.

```{r}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, eval = FALSE)
```

```{r}
library(divest)
library(doFuture)
library(dplyr)
library(foreach)
library(fslr)
library(oro.dicom)
library(oro.nifti)
library(neurobase)
library(progressr)
library(purrr)
library(readr)
library(stringr)
library(tibble)
library(tidyverse)
library(lubridate)
handlers(global = TRUE)
```

12 images produced errors during preprocessing and were skipped. The IDs of these images are listed in the chunk below.

```{r}
skip_imgs <- c(
  "I436253",
  "I59843",
  "I272379",
  "I135611",
  "I48503",
  "I77171",
  "I154113",
  "I74064",
  "I32485",
  "I298204",
  "I436281",
  "I35210"
)

img_dirs <- list.files("data/ADNI", recursive = TRUE, full.names = TRUE) %>%
  gsub(pattern = "([^/]+$)", replacement = "") %>%
  unique()
```

We start with T1-weighted images in DICOM format, with the files for each image being contined within a directory, contained within the `img_dirs` vector above. The `fslr` package is used to interface with FSL to process the images. We also consider study information obtained directly from the ADNI online repository, providing details regarding participants' demographics, study group, and imaging history. 

```{r}
adni_info <- read_csv("data/adni_info.csv")
adni_info <- adni_info |>
  rename(
    image_id = "Image Data ID",
    subid = "Subject",
    date = "Acq Date"
  ) |>
  mutate(
    date = mdy(date)
  ) |>
  select(
    subid,
    Group,
    Sex,
    Age,
    image_id,
    date
  ) |>
  arrange(subid, date)
```

Each image is segmented using FSL's `run_first_all` function. This command wraps several lower-level processing steps, including linear image registration to the MNI space. It then segments the following structures: putamen, caudate nucleus, nucleus accumbens, globus pallidus, hippocampus, amygdala, thalamus, brain stem. These segmentations are loaded, with the `fslstats` function used to compute the FSL-estimated volumes for the relevant structures (left and right hippocampus and left and right thalamus). Finally, for each structure, the structure estimate taken from the segmentation is loaded, and the indices of all voxels with non-zero intensity values are identified. The voxels found to be extreme in at least one dimension are taken to be voxels that lie along the structure's surface, and the indices of these voxels are saved as the x, y, and z coordinates in 3-dimensional space. For each structure, a matrix was created including the coordinates of all surface voxels for each image.

```{r}
plan(sequential)

image_info <- foreach(
  img_idx = seq_along(img_dirs),
  .errorhandling = "pass"
) %do%
  {
    img_val <- str_split(img_dirs[img_idx], "/")[[1]][6]
    print(paste0("Image ", img_idx, " of ", length(img_dirs), ": ", img_val))

    # check if image has already been segmented - if so, skip to the next image
    proc_dir <- paste0(
      "data/adni_processed_fsl",
      gsub(pattern = "data/ADNI", replacement = "", img_dirs[img_idx])
    )
    if (file.exists(paste0(proc_dir, "_all_fast_origsegs.nii.gz"))) {
      print("Already segmented...")
      NULL
    } else if (img_val %in% skip_imgs) {
      print("Skipping...")
      NULL
    } else {
      print("Segmenting...")
      nii <- readDicom(img_dirs[img_idx], interactive = FALSE, verbosity = -1)

      # segment image and save results to drive
      run_first_all(
        nii,
        oprefix = proc_dir,
        verbose = FALSE,
        opts = "-d"
      )
      gc()
    }

    if (file.exists(paste0(proc_dir, "_all_fast_origsegs.nii.gz"))) {
      print("Getting Scan Information...")
      temp_segs <- readnii(paste0(proc_dir, "_all_fast_origsegs.nii.gz"))
      seg_vols <- fslstats(
        paste0(proc_dir, "_all_fast_origsegs.nii.gz"),
        opts = "-V",
        ts = TRUE
      )

      # left and right hippocampi correspond with indices 6 and 13, respectively
      lhipp_vol <- seg_vols[6] %>%
        str_replace(
          pattern = "([^\\s]+)",
          replacement = ""
        ) %>%
        gsub(
          pattern = " ",
          replacement = ""
        ) %>%
        as.numeric()

      rhipp_vol <- seg_vols[13] %>%
        str_replace(
          pattern = "([^\\s]+)",
          replacement = ""
        ) %>%
        gsub(
          pattern = " ",
          replacement = ""
        ) %>%
        as.numeric()

      # left and right thalami correspond with indices 1 and 9, respectively
      lthal_vol <- seg_vols[1] %>%
        str_replace(
          pattern = "([^\\s]+)",
          replacement = ""
        ) %>%
        gsub(
          pattern = " ",
          replacement = ""
        ) %>%
        as.numeric()

      rthal_vol <- seg_vols[9] %>%
        str_replace(
          pattern = "([^\\s]+)",
          replacement = ""
        ) %>%
        gsub(
          pattern = " ",
          replacement = ""
        ) %>%
        as.numeric()

      scan_info <- tibble(
        image_id = img_val,
        lhipp_vol = lhipp_vol,
        rhipp_vol = rhipp_vol,
        lthal_vol = lthal_vol,
        rthal_vol = rthal_vol
      )

      if (file.exists(paste0(proc_dir, "lhipp.csv"))) {
        NULL
      } else if (img_val %in% skip_imgs) {
        NULL
      } else {
        print("Extracting LHIPP Surface...")
        lhipp <- temp_segs[,,, 6]

        # In future, move to function
        # find indices of all voxels with intensity > 0
        lhipp_idx <- which(lhipp > 0, arr.ind = TRUE)
        surface <- rep(FALSE, nrow(lhipp_idx))
        for (dim in 1:3) {
          for (dim2 in 1:3) {
            if (dim != dim2) {
              dim3 <- seq(1, 3, 1)[!(1:3 %in% c(dim, dim2))]
              # at each point in a 2-dimensional profile view of structure,
              # obtain all "depth" coordinates of non-zero intensity voxels
              unique_vals <- unique(lhipp_idx[, c(dim, dim2)])
              for (i in seq_len(nrow(unique_vals))) {
                vals <- lhipp_idx[
                  lhipp_idx[, dim] == unique_vals[i, 1] &
                    lhipp_idx[, dim2] == unique_vals[i, 2],
                  dim3
                ]
                min_vals <- vector()
                max_vals <- vector()
                # voxel is assumed to be a surface point if there are no
                # neighboring voxels in at least one direction along dim
                for (v in vals) {
                  if (!((v - 1) %in% vals)) {
                    min_vals <- c(min_vals, v)
                  } else if (!((v + 1) %in% vals)) {
                    max_vals <- c(max_vals, v)
                  }
                }
                surface_vals <- c(min_vals, max_vals)
                for (v in surface_vals) {
                  surface[which(
                    lhipp_idx[, dim] == unique_vals[i, 1] &
                      lhipp_idx[, dim2] == unique_vals[i, 2] &
                      lhipp_idx[, dim3] == v,
                    arr.ind = TRUE
                  )] <- TRUE
                }
              }
            } else {
              next
            }
          }
        }
        lhipp_mat <- cbind(
          img_val,
          lhipp_idx %*% diag(pixdim(as.nifti(lhipp))[2:4]),
          lhipp[lhipp_idx]
        )
        df_names <- c("scan_id", "x", "y", "z", "intensity")
        lhipp_df <- as_tibble(lhipp_mat, .name_repair = "minimal")
        names(lhipp_df) <- df_names
        lhipp_surface <- cbind(
          img_val,
          lhipp_idx[surface, ] %*% diag(pixdim(as.nifti(lhipp))[2:4]),
          lhipp[lhipp_idx[surface, ]]
        )
        lhipp_surface_df <- as_tibble(
          lhipp_surface,
          .name_repair = "minimal"
        )
        names(lhipp_surface_df) <- df_names

        write_csv(lhipp_df, paste0(proc_dir, "lhipp.csv"))
        write_csv(lhipp_surface_df, paste0(proc_dir, "lhipp_surface.csv"))
      }

      # repeating process above for other structures of interest
      if (file.exists(paste0(proc_dir, "rhipp.csv"))) {
        NULL
      } else if (img_val %in% skip_imgs) {
        NULL
      } else {
        print("Extracting RHIPP Surface...")
        rhipp <- temp_segs[,,, 13]
        rhipp_idx <- which(rhipp > 0, arr.ind = TRUE)
        surface <- rep(FALSE, nrow(rhipp_idx))
        for (dim in 1:3) {
          for (dim2 in 1:3) {
            if (dim != dim2) {
              dim3 <- seq(1, 3, 1)[!(1:3 %in% c(dim, dim2))]
              unique_vals <- unique(rhipp_idx[, c(dim, dim2)])
              for (i in seq_len(nrow(unique_vals))) {
                vals <- rhipp_idx[
                  rhipp_idx[, dim] == unique_vals[i, 1] &
                    rhipp_idx[, dim2] == unique_vals[i, 2],
                  dim3
                ]
                min_vals <- vector()
                max_vals <- vector()
                for (v in vals) {
                  if (!((v - 1) %in% vals)) {
                    min_vals <- c(min_vals, v)
                  } else if (!((v + 1) %in% vals)) {
                    max_vals <- c(max_vals, v)
                  }
                }
                surface_vals <- c(min_vals, max_vals)
                for (v in surface_vals) {
                  surface[which(
                    rhipp_idx[, dim] == unique_vals[i, 1] &
                      rhipp_idx[, dim2] == unique_vals[i, 2] &
                      rhipp_idx[, dim3] == v,
                    arr.ind = TRUE
                  )] <- TRUE
                }
              }
            } else {
              next
            }
          }
        }
        rhipp_mat <- cbind(
          img_val,
          rhipp_idx %*% diag(pixdim(as.nifti(rhipp))[2:4]),
          rhipp[rhipp_idx]
        )
        df_names <- c("scan_id", "x", "y", "z", "intensity")
        rhipp_df <- as_tibble(rhipp_mat, .name_repair = "minimal")
        names(rhipp_df) <- df_names
        rhipp_surface <- cbind(
          img_val,
          rhipp_idx[surface, ] %*% diag(pixdim(as.nifti(rhipp))[2:4]),
          rhipp[rhipp_idx[surface, ]]
        )
        rhipp_surface_df <- as_tibble(
          rhipp_surface,
          .name_repair = "minimal"
        )
        names(rhipp_surface_df) <- df_names

        write_csv(rhipp_df, paste0(proc_dir, "rhipp.csv"))
        write_csv(rhipp_surface_df, paste0(proc_dir, "rhipp_surface.csv"))
      }

      if (file.exists(paste0(proc_dir, "lthal.csv"))) {
        NULL
      } else if (img_val %in% skip_imgs) {
        NULL
      } else {
        print("Extracting LTHAL Surface...")
        lthal <- temp_segs[,,, 1]
        lthal_idx <- which(lthal > 0, arr.ind = TRUE)
        surface <- rep(FALSE, nrow(lthal_idx))
        for (dim in 1:3) {
          for (dim2 in 1:3) {
            if (dim != dim2) {
              dim3 <- seq(1, 3, 1)[!(1:3 %in% c(dim, dim2))]
              unique_vals <- unique(lthal_idx[, c(dim, dim2)])
              for (i in seq_len(nrow(unique_vals))) {
                vals <- lthal_idx[
                  lthal_idx[, dim] == unique_vals[i, 1] &
                    lthal_idx[, dim2] == unique_vals[i, 2],
                  dim3
                ]
                min_vals <- vector()
                max_vals <- vector()
                for (v in vals) {
                  if (!((v - 1) %in% vals)) {
                    min_vals <- c(min_vals, v)
                  } else if (!((v + 1) %in% vals)) {
                    max_vals <- c(max_vals, v)
                  }
                }
                surface_vals <- c(min_vals, max_vals)
                for (v in surface_vals) {
                  surface[which(
                    lthal_idx[, dim] == unique_vals[i, 1] &
                      lthal_idx[, dim2] == unique_vals[i, 2] &
                      lthal_idx[, dim3] == v,
                    arr.ind = TRUE
                  )] <- TRUE
                }
              }
            } else {
              next
            }
          }
        }
        lthal_mat <- cbind(
          img_val,
          lthal_idx %*% diag(pixdim(as.nifti(lthal))[2:4]),
          lthal[lthal_idx]
        )
        df_names <- c("scan_id", "x", "y", "z", "intensity")
        lthal_df <- as_tibble(lthal_mat, .name_repair = "minimal")
        names(lthal_df) <- df_names
        lthal_surface <- cbind(
          img_val,
          lthal_idx[surface, ] %*% diag(pixdim(as.nifti(lthal))[2:4]),
          lthal[lthal_idx[surface, ]]
        )
        lthal_surface_df <- as_tibble(
          lthal_surface,
          .name_repair = "minimal"
        )
        names(lthal_surface_df) <- df_names

        write_csv(lthal_df, paste0(proc_dir, "lthal.csv"))
        write_csv(lthal_surface_df, paste0(proc_dir, "lthal_surface.csv"))
      }

      if (file.exists(paste0(proc_dir, "rthal.csv"))) {
        NULL
      } else if (img_val %in% skip_imgs) {
        NULL
      } else {
        print("Extracting RTHAL Surface...")
        rthal <- temp_segs[,,, 9]
        rthal_idx <- which(rthal > 0, arr.ind = TRUE)
        surface <- rep(FALSE, nrow(rthal_idx))
        for (dim in 1:3) {
          for (dim2 in 1:3) {
            if (dim != dim2) {
              dim3 <- seq(1, 3, 1)[!(1:3 %in% c(dim, dim2))]
              unique_vals <- unique(rthal_idx[, c(dim, dim2)])
              for (i in seq_len(nrow(unique_vals))) {
                vals <- rthal_idx[
                  rthal_idx[, dim] == unique_vals[i, 1] &
                    rthal_idx[, dim2] == unique_vals[i, 2],
                  dim3
                ]
                min_vals <- vector()
                max_vals <- vector()
                for (v in vals) {
                  if (!((v - 1) %in% vals)) {
                    min_vals <- c(min_vals, v)
                  } else if (!((v + 1) %in% vals)) {
                    max_vals <- c(max_vals, v)
                  }
                }
                surface_vals <- c(min_vals, max_vals)
                for (v in surface_vals) {
                  surface[which(
                    rthal_idx[, dim] == unique_vals[i, 1] &
                      rthal_idx[, dim2] == unique_vals[i, 2] &
                      rthal_idx[, dim3] == v,
                    arr.ind = TRUE
                  )] <- TRUE
                }
              }
            } else {
              next
            }
          }
        }
        rthal_mat <- cbind(
          img_val,
          rthal_idx %*% diag(pixdim(as.nifti(rthal))[2:4]),
          rthal[rthal_idx]
        )
        df_names <- c("scan_id", "x", "y", "z", "intensity")
        rthal_df <- as_tibble(rthal_mat, .name_repair = "minimal")
        names(rthal_df) <- df_names
        rthal_surface <- cbind(
          img_val,
          rthal_idx[surface, ] %*% diag(pixdim(as.nifti(rthal))[2:4]),
          rthal[rthal_idx[surface, ]]
        )
        rthal_surface_df <- as_tibble(
          rthal_surface,
          .name_repair = "minimal"
        )
        names(rthal_surface_df) <- df_names

        write_csv(rthal_df, paste0(proc_dir, "rthal.csv"))
        write_csv(rthal_surface_df, paste0(proc_dir, "rthal_surface.csv"))
      }
    }
    scan_info
  }
```

After extracting all surface voxels, the chunk above returns a list of `tibble` objects, with each containing the FSL-estimated volumes for the structures of interest. This information is joined to the study information and saved below. The following chunk also saves information about segmentation errors that were encountered.

```{r}
info_present <- map(
  image_info,
  ~ !(is.null(.x) | "simpleError" %in% class(.x))
) |>
  reduce(c)
image_info <- image_info[info_present] |>
  reduce(bind_rows)


adni_info <- full_join(
  adni_info,
  image_info,
  by = "image_id"
)

write_csv(adni_info, "data/adni_info_full.csv")

processed_dirs <- gsub(
  pattern = "data/ADNI",
  replacement = "data/adni_processed_fsl",
  img_dirs
)
error_list <- map(
  processed_dirs,
  ~ ifelse(
    file.exists(paste0(.x, "/_all_fast_origsegs.nii.gz")),
    FALSE,
    TRUE
  )
) |>
  reduce(c)
segmentation_errors <- img_dirs[error_list]

saveRDS(segmentation_errors, "data/segmentation_errors.rds")
```
