---
title: "ADNI Application Code"
author: "Robert Zielinski"
date: "December 10, 2025"
format: html
---

This notebook contains the code necessary to run the PME and LPME algorithms on the processed MRI data from the ADNI study. The code assumes that the data for the left and right hippocampi and left and right thalamuses of all participants of interest are stored in the following files:

- `data/lhipp_surface_fsl.csv`
- `data/rhipp_surface_fsl.csv`
- `data/lthal_surface_fsl.csv`
- `data/rthal_surface_fsl.csv`

The previous notebook, `code/03_adni_mri_preprocess.qmd`, includes the code used to preprocess the MRI data into the required format.

```{r}
knitr::opts_chunk$set(eval = FALSE)
```

```{r}
library(reticulate)
use_condaenv("lpme")

library(here)
library(lubridate)
library(future)
library(plotly)
library(pme)
library(pracma)
library(RColorBrewer)
library(Rfast)
library(tictoc)
library(tidyverse)

source(here("code/functions/preprocess_adni.R"))
source(here("code/functions/fit_adni.R"))

options(rgl.useNULL = TRUE)
```

The `preprocess_adni` function takes in a matrix of surface voxel data and applies preprocessing steps to prepare the data for analysis. First, images with the number of surface voxels falling below the 0.25th percentile are assumed to have faulty segmentations and are excluded. Then, the voxel coordinates are centered and scaled to fall between -1 and 1 in each dimension. The voxel coordinates are augmented with spherical coordinates, though the results presented in the paper relied on partitioning instead. The collection of image data was filtered to include only participants who meet minimum follow-up duration and number of visit thresholds.

```{r}
preprocess_adni <- function(
  adni_surface,
  adni_info,
  min_duration = 2,
  min_visits = 3,
  scan_exclude_threshold = 0.0025
) {
  require(dplyr)
  require(lubridate)
  require(tidyr)

  adni_surface <- adni_surface |>
    left_join(adni_info, by = join_by(scan_id == image_id))

  adni_surface <- adni_surface |>
    mutate(date = decimal_date(date))

  adni_n <- adni_surface |>
    group_by(subid, scan_id) |>
    tally() |>
    ungroup()
  adni_threshold <- quantile(adni_n$n, scan_exclude_threshold)

  exclude_scans <- adni_n |>
    filter(n < adni_threshold) |>
    select(scan_id) |>
    unlist()

  adni_surface <- adni_surface |>
    filter(!(scan_id %in% exclude_scans))

  adni_centers <- adni_surface |>
    group_by(subid, date, scan_id) |>
    summarise(
      mean_x = mean(x),
      mean_y = mean(y),
      mean_z = mean(z),
      max_x = max(abs(x - mean_x)),
      max_y = max(abs(y - mean_y)),
      max_z = max(abs(z - mean_z))
    ) |>
    ungroup()

  adni_bl <- adni_surface |>
    group_by(subid) |>
    arrange(date) |>
    summarize(
      date_bl = first(date),
      date_max = max(date),
      n_visits = length(unique(date))
    ) |>
    mutate(duration = date_max - date_bl)

  adni_surface <- adni_surface |>
    left_join(adni_centers, by = c("subid", "date", "scan_id")) |>
    left_join(adni_bl, by = "subid") |>
    mutate(
      x = (x - mean_x) / max_x,
      y = (y - mean_y) / max_y,
      z = (z - mean_z) / max_z,
      time_from_bl = (date - date_bl) / duration
    ) |>
    filter(duration > min_duration, n_visits >= min_visits)

  adni_surface_spherical <- adni_surface |>
    select(x, y, z) |>
    as.matrix() |>
    cart2sph() |>
    as_tibble()

  adni_surface <- bind_cols(adni_surface, adni_surface_spherical)

  list(surface = adni_surface, centers = adni_centers)
}
```


The `fit_adni` function takes the preprocessed data output by `preprocess_adni()` and loops through the images for each participant, applying the relevant manifold estimation methods for each individual. For each participant, the surface data are partitioned using the $xy$-plane. Then, the following models are fit:

- Augmented LPME
- Partitioned LPME
- Augmented PME
- Partitioned PME
- Partitioned Principal Surface

After model fitting, the volumes and profile surface areas of the estimated structures are calculated as one indication of the stability of the structure estimates. While the volume estimation approach yielded insufficient estimates, comparisons of the surface areas of the manifold estimates projected to 2-dimensional planes proved to appropriately represent the results in most cases. Specifically, the `mesh_projection` function found in `code/functions/mesh_projection.R` is used to generate these areas by projecting the estimated manifold reconstructions to each of the three coordinate planes in 3-dimensional Euclidean space, then calculating the area of these projections using a 2-dimensional Delaunay triangulation. These results are then saved for each individual and structure.

```{r}
fit_adni <- function(
  adni_surface,
  adni_centers,
  structure,
  partition_overlap = 0,
  cores = parallelly::availableCores(),
  verbose = FALSE
) {
  require(dplyr)
  require(future)
  require(future.batchtools)
  require(future.mirai)
  require(here)
  require(mirai)
  require(parallel)
  require(pme)
  require(purrr)
  require(tidyr)
  require(progressr)

  require(reticulate)
  use_condaenv("lpme")

  require(doRNG)
  require(doSNOW)
  require(foreach)

  source(here("code/functions/calculate_lpme_reconstructions.R"))
  source(here("code/functions/calculate_pme_reconstructions.R"))
  source(here("code/functions/estimate_volume.R"))
  source(here("code/functions/interior_identification.R"))
  source(here("code/prinSurf_v3.R"))
  source(here("code/functions/estimate_volume_interior.R"))
  source(here("code/functions/estimate_mesh_volume_poisson.R"))
  source(here("code/functions/mesh_projection.R"))

  handlers(global = TRUE)

  patnos <- unique(adni_surface$subid)

  # do not repeat fitting when estimates already exist for this participant
  remove_patnos <- map(
    patnos,
    ~ !file.exists(here(paste0(
      "output/adni_area/",
      .x,
      "/",
      structure,
      "_results.RDS"
    )))
  ) |>
    reduce(c)

  patnos <- patnos[remove_patnos]

  # parameters for 3-dimensional mesh reconstruction should differ by structure
  # this approach is not used in results
  if (grepl("hipp", structure) == TRUE) {
    alpha_vals <- exp(seq(-3, 0, 0.25))
  } else {
    alpha_vals <- exp(seq(-2, 0, 0.25))
  }

  # parallel processing setup, if applicable
  if (verbose == FALSE) {
    require(tictoc)
    require(pme)
    require(dplyr)
    require(tidyr)

    cl <- makeCluster(cores, outfile = "")
    registerDoSNOW(cl)
  } else {
    plan(sequential, split = TRUE)
    require(tictoc)
    cl <- makeCluster(1, outfile = "")
    registerDoSNOW(cl)
  }

  set.seed(500)

  pb <- txtProgressBar(max = length(patnos), style = 3)
  progress <- function(n) setTxtProgressBar(pb, n)
  opts <- list(progress = progress)

  registerDoRNG(716241)

  # loop through participants in parallel
  # adni_fit is not used in practice; results are saved as they are generated
  adni_fit <- foreach(
    patno_idx = seq_along(patnos),
    .export = c(
      "patnos",
      "alpha_vals",
      "adni_surface",
      "adni_centers",
      "structure",
      "partition_overlap",
      "verbose",
      "prinSurf",
      "calculate_lpme_reconstructions",
      "calculate_pme_reconstructions",
      "estimate_volume_interior_lpme",
      "estimate_volume_interior_pme",
      "interior_identification",
      "get_orientation",
      "estimate_mesh_volume_poisson",
      "mesh_projection"
    ),
    .packages = c(
      "alphashape3d",
      "dplyr",
      "here",
      "lubridate",
      "plotly",
      "plot3D",
      "pme",
      "pracma",
      "purrr",
      "RColorBrewer",
      "Rfast",
      "tictoc",
      "tidyr",
      "reticulate"
    ),
    .inorder = TRUE,
    .options.snow = opts,
    .errorhandling = "stop"
  ) %dopar%
    {
      use_condaenv("lpme")
      patno <- patnos[patno_idx]

      # create partitioning indicators
      # allow for slight overlap in partitions to avoid gaps in final structure
      patno_adni <- adni_surface |>
        filter(subid == patno) |>
        select(time_from_bl, x, y, z, theta, phi) |>
        mutate(
          partition = z > 0,
          partition1 = z > -partition_overlap,
          partition2 = z < partition_overlap
        )

      patno_adni_centers <- adni_centers |>
        filter(subid == patno) |>
        group_by(date) |>
        summarize(
          max_x = max(max_x),
          max_y = max(max_y),
          max_z = max(max_z)
        )

      adni_aug <- patno_adni |>
        select(-partition, -partition1, -partition2) |>
        as.matrix()

      adni_pt1 <- patno_adni |>
        filter(partition1 == TRUE) |>
        select(time_from_bl, x, y, z) |>
        as.matrix()

      adni_pt2 <- patno_adni |>
        filter(partition2 == TRUE) |>
        select(time_from_bl, x, y, z) |>
        as.matrix()

      # compute minimum number of observations across time points
      # used for LPME initialization - can't have more cluster components than
      # surface voxels
      min_obs_aug <- patno_adni |>
        group_by(time_from_bl) |>
        summarize(n = n()) |>
        select(n) |>
        min()

      min_obs_pt1 <- patno_adni |>
        filter(partition1 == TRUE) |>
        group_by(time_from_bl) |>
        summarize(n = n()) |>
        select(n) |>
        min()

      min_obs_pt2 <- patno_adni |>
        filter(partition2 == TRUE) |>
        group_by(time_from_bl) |>
        summarize(n = n()) |>
        select(n) |>
        min()

      time_values <- unique(patno_adni$time_from_bl)

      adni_pme_aug_list <- list()
      adni_pme_aug_time_list <- list()
      adni_pme_aug_reconstruction_list <- list()

      # create empty lists to store model results
      adni_lpme_aug <- NULL
      adni_lpme_part <- replicate(2, NULL, simplify = FALSE)
      adni_pme_part <- replicate(
        2,
        replicate(length(time_values), NULL, simplify = FALSE),
        simplify = FALSE
      )
      adni_pc_part <- replicate(
        2,
        replicate(length(time_values), NULL, simplify = FALSE),
        simplify = FALSE
      )

      adni_lpme_aug_reconstructions <- NULL
      adni_lpme_part_reconstructions <- replicate(
        2,
        NULL,
        simplify = FALSE
      )
      adni_pme_part_reconstructions <- replicate(
        2,
        replicate(length(time_values), NULL, simplify = FALSE),
        simplify = FALSE
      )
      adni_pc_part_reconstructions <- replicate(
        2,
        replicate(length(time_values), NULL, simplify = FALSE),
        simplify = FALSE
      )

      adni_lpme_aug_time <- NULL
      adni_lpme_part_times <- replicate(2, NULL, simplify = FALSE)
      adni_pme_part_times <- replicate(
        2,
        replicate(length(time_values), NULL, simplify = FALSE),
        simplify = FALSE
      )
      adni_pc_part_times <- replicate(
        2,
        replicate(length(time_values), NULL, simplify = FALSE),
        simplify = FALSE
      )

      if (verbose) {
        print(paste0("Fitting augmented LPME"))
      }

      tryCatch(
        {
          tic()
          adni_lpme_aug <- lpme(
            adni_aug,
            d = 2,
            min_clusters = min(20, floor(min_obs_aug / 2)),
            verbose = verbose,
            print_plots = FALSE
          )
          adni_lpme_aug_time_end <- toc(quiet = TRUE)
          adni_lpme_aug_time <- adni_lpme_aug_time_end$toc -
            adni_lpme_aug_time_end$tic

          adni_lpme_aug_reconstructions <- calculate_lpme_reconstructions(
            adni_lpme_aug,
            adni_aug
          )
        },
        error = function(e) {
          print(
            paste0("Augmented LPME failed for patno ", patno, ":"),
            e$message
          )
        }
      )

      if (verbose) {
        print(paste0("Fitting partitioned LPME"))
      }

      tryCatch(
        {
          tic()

          adni_lpme_part[[1]] <- lpme(
            adni_pt1,
            d = 2,
            min_clusters = min(20, floor(min_obs_pt1 - 2)),
            verbose = verbose,
            print_plots = FALSE
          )

          adni_lpme_time_pt1 <- toc(quiet = TRUE)
          adni_lpme_part_times[[1]] <- adni_lpme_time_pt1$toc -
            adni_lpme_time_pt1$tic
          adni_lpme_part_reconstructions[[
            1
          ]] <- calculate_lpme_reconstructions(
            adni_lpme_part[[1]],
            adni_pt1[adni_pt1[, 4] > 0, ]
          )
        },
        error = function(e) {
          print(paste0("LPME Part 1 failed for patno ", patno, ": "), e$message)
        }
      )

      tryCatch(
        {
          tic()
          adni_lpme_part[[2]] <- lpme(
            adni_pt2,
            d = 2,
            min_clusters = min(20, floor(min_obs_pt2 - 2)),
            verbose = verbose,
            print_plots = FALSE
          )
          adni_lpme_time_pt2 <- toc(quiet = TRUE)
          adni_lpme_part_times[[2]] <- adni_lpme_time_pt2$toc -
            adni_lpme_time_pt2$tic
          adni_lpme_part_reconstructions[[
            2
          ]] <- calculate_lpme_reconstructions(
            adni_lpme_part[[2]],
            adni_pt2[adni_pt2[, 4] <= 0, ]
          )
        },
        error = function(e) {
          print(paste0("LPME Part 2 failed for patno ", patno, ": "), e$message)
        }
      )

      if (verbose) {
        print(paste0("Fitting time point-specific estimates"))
      }
      for (time_idx in seq_along(time_values)) {
        temp_adni <- adni_aug[adni_aug[, 1] == time_values[time_idx], -1]
        temp_adni_pt1 <- adni_pt1[
          adni_pt1[, 1] == time_values[time_idx],
          -1
        ]
        temp_adni_pt2 <- adni_pt2[
          adni_pt2[, 1] == time_values[time_idx],
          -1
        ]

        tryCatch(
          {
            tic()
            adni_pme_aug_list[[time_idx]] <- pme(
              temp_adni,
              d = 2,
              min_clusters = min(20, floor(nrow(temp_adni) / 2)),
              verbose = FALSE
            )
            temp_pme_aug_time <- toc(quiet = TRUE)
            adni_pme_aug_time_list[[time_idx]] <- temp_pme_aug_time$toc -
              temp_pme_aug_time$tic

            temp_adni_pme_aug_reconstructions <- calculate_pme_reconstructions(
              adni_pme_aug_list[[time_idx]],
              temp_adni
            )
            adni_pme_aug_reconstruction_list[[time_idx]] <- cbind(
              time_values[time_idx],
              temp_adni_pme_aug_reconstructions
            )
          },
          error = function(e) {
            print(
              paste0(
                "Augmented PME failed at time point ",
                time_idx,
                " for patno ",
                patno,
                ": "
              ),
              e$message
            )
          }
        )

        tryCatch(
          {
            tic()
            adni_pme_part[[1]][[time_idx]] <- pme(
              temp_adni_pt1,
              d = 2,
              min_clusters = min(20, floor(nrow(temp_adni_pt1) / 2))
            )
            temp_pme_time_pt1 <- toc(quiet = TRUE)
            adni_pme_part_times[[1]][[time_idx]] <- temp_pme_time_pt1$toc -
              temp_pme_time_pt1$tic
            temp_adni_pme_reconstructions_pt1 <- calculate_pme_reconstructions(
              adni_pme_part[[1]][[time_idx]],
              temp_adni_pt1[temp_adni_pt1[, 3] > 0, ]
            )
            adni_pme_part_reconstructions[[1]][[time_idx]] <- cbind(
              time_values[time_idx],
              temp_adni_pme_reconstructions_pt1
            )
          },
          error = function(e) {
            print(
              paste0(
                "PME Part 1 failed at time point ",
                time_idx,
                " for patno ",
                patno,
                ": "
              ),
              e$message
            )
          }
        )

        tryCatch(
          {
            tic()
            adni_pme_part[[2]][[time_idx]] <- pme(
              temp_adni_pt2,
              d = 2,
              min_clusters = min(20, floor(nrow(temp_adni_pt2) / 2))
            )
            temp_pme_time_pt2 <- toc(quiet = TRUE)
            adni_pme_part_times[[2]][[time_idx]] <- temp_pme_time_pt2$toc -
              temp_pme_time_pt2$tic
            temp_adni_pme_reconstructions_pt2 <- calculate_pme_reconstructions(
              adni_pme_part[[2]][[time_idx]],
              temp_adni_pt2[temp_adni_pt2[, 3] <= 0, ]
            )
            adni_pme_part_reconstructions[[2]][[time_idx]] <- cbind(
              time_values[time_idx],
              temp_adni_pme_reconstructions_pt2
            )
          },
          error = function(e) {
            print(
              paste0(
                "PME Part 2 failed at time point ",
                time_idx,
                " for patno ",
                patno,
                ": "
              ),
              e$message
            )
          }
        )

        tryCatch(
          {
            tic()
            principal_surface_part1 <- prinSurf(temp_adni_pt1, flag = FALSE)
            temp_pc_time_pt1 <- toc(quiet = TRUE)
            adni_pc_part_times[[1]][[time_idx]] <- temp_pc_time_pt1$toc -
              temp_pc_time_pt1$tic
            surface_mse_part1 <- map(
              seq_along(principal_surface_part1),
              ~ principal_surface_part1[[.x]]$MSE
            ) |>
              unlist()
            opt_surface_part1 <- which.min(surface_mse_part1)

            adni_pc_part[[1]][[time_idx]] <- principal_surface_part1[[
              opt_surface_part1 + 2
            ]]
            temp_adni_pc_reconstructions_pt1 <- cbind(
              time_values[time_idx],
              principal_surface_part1[[opt_surface_part1 + 2]]$PS
            )
            adni_pc_part_reconstructions[[1]][[
              time_idx
            ]] <- temp_adni_pc_reconstructions_pt1[temp_adni_pt1[, 3] > 0, ]
          },
          error = function(e) {
            print(
              paste0(
                "PS Part 1 failed at time point ",
                time_idx,
                " for patno ",
                patno,
                ": "
              ),
              e$message
            )
          }
        )

        tryCatch(
          {
            tic()
            principal_surface_part2 <- prinSurf(temp_adni_pt2, flag = FALSE)
            temp_pc_time_pt2 <- toc(quiet = TRUE)
            adni_pc_part_times[[2]][[time_idx]] <- temp_pc_time_pt2$toc -
              temp_pc_time_pt2$tic
            surface_mse_part2 <- map(
              seq_along(principal_surface_part2),
              ~ principal_surface_part2[[.x]]$MSE
            ) |>
              unlist()
            opt_surface_part2 <- which.min(surface_mse_part2)

            adni_pc_part[[2]][[time_idx]] <- principal_surface_part2[[
              opt_surface_part2 + 2
            ]]
            temp_adni_pc_reconstructions_pt2 <- cbind(
              time_values[time_idx],
              principal_surface_part2[[opt_surface_part2 + 2]]$PS
            )
            adni_pc_part_reconstructions[[2]][[
              time_idx
            ]] <- temp_adni_pc_reconstructions_pt2[temp_adni_pt2[, 3] <= 0, ]
          },
          error = function(e) {
            print(
              paste0(
                "PS Part 2 failed at time point ",
                time_idx,
                " for patno ",
                patno,
                ": ",
              ),
              e$message
            )
          }
        )
      }

      if (verbose) {
        print(paste0("Fitting complete"))
      }

      # collapse lists of manifold estimate reconstructions and compute times
      # compute times for partitioned models are summed
      if (!(0 %in% lengths(adni_lpme_part_reconstructions))) {
        adni_lpme_part_reconstructions <- reduce(
          adni_lpme_part_reconstructions,
          rbind
        )
        adni_lpme_part_times <- reduce(adni_lpme_part_times, sum)
      } else {
        adni_lpme_part_reconstructions <- NULL
        adni_lpme_part_times <- NULL
      }

      adni_pme_aug_reconstructions <- NULL
      if (!(0 %in% lengths(adni_pme_aug_reconstruction_list))) {
        adni_pme_aug_reconstructions <- reduce(
          adni_pme_aug_reconstruction_list,
          rbind
        )
      }

      adni_pme_aug_time_list <- ifelse(
        !(0 %in% lengths(adni_pme_aug_time_list)),
        reduce(
          adni_pme_aug_time_list,
          sum
        ),
        NULL
      )

      adni_pme_part_complete <- map(
        seq_along(adni_pme_part_reconstructions),
        ~ !(0 %in% lengths(adni_pme_part_reconstructions[[.x]]))
      ) |>
        reduce(c) |>
        sum()

      if (adni_pme_part_complete == length(adni_pme_part_reconstructions)) {
        adni_pme_part_reconstructions <- map(
          seq_along(adni_pme_part_reconstructions),
          ~ reduce(adni_pme_part_reconstructions[[.x]], rbind)
        ) |>
          reduce(rbind)

        adni_pme_part_times <- map(
          seq_along(adni_pme_part_times),
          ~ reduce(adni_pme_part_times[[.x]], sum)
        ) |>
          reduce(sum)
      }

      adni_pc_part_complete <- map(
        seq_along(adni_pc_part_reconstructions),
        ~ !(0 %in% lengths(adni_pc_part_reconstructions[[.x]]))
      ) |>
        reduce(c) |>
        sum()

      if (adni_pc_part_complete == length(adni_pc_part_reconstructions)) {
        adni_pc_part_reconstructions <- map(
          seq_along(adni_pc_part_reconstructions),
          ~ reduce(adni_pc_part_reconstructions[[.x]], rbind)
        ) |>
          reduce(rbind)

        adni_pc_part_times <- map(
          seq_along(adni_pc_part_times),
          ~ reduce(adni_pc_part_times[[.x]], sum)
        ) |>
          reduce(sum)
      }

      if (verbose) {
        print(paste0("Estimating volumes"))
      }

      if (!(0 %in% lengths(adni_lpme_part))) {
        adni_lpme_part_volumes <- tryCatch(
          {
            adni_lpme_part_interior_volume <- estimate_volume_interior_lpme(
              adni_lpme_part,
              list(adni_pt1, adni_pt2),
              time_values,
              n_points = 10000,
              data_max = patno_adni_centers,
              limit_scaler = 0.05,
              partition_index = 3
            )
            adni_lpme_part_interior_volume$volumes
          },
          error = function(e) {
            print(
              paste0(
                "Error in LPME Part interior volume calculation for patno ",
                patno,
                ": "
              ),
              e$message
            )
            NULL
          }
        )
      } else {
        adni_lpme_part_volumes <- NULL
      }

      pme_part_complete <- map(
        seq_along(adni_pme_part),
        ~ !(0 %in% lengths(adni_pme_part[[.x]]))
      ) |>
        reduce(c) |>
        sum()

      if (pme_part_complete == length(adni_pme_part)) {
        adni_pme_part_volumes <- tryCatch(
          {
            adni_pme_part_interior_volumes <- estimate_volume_interior_pme(
              adni_pme_part,
              list(adni_pt1, adni_pt2),
              time_values,
              n_points = 10000,
              data_max = patno_adni_centers,
              limit_scaler = 0.05,
              partition_index = 3
            )
            adni_pme_part_interior_volumes$volumes
          },
          error = function(e) {
            print(
              paste0(
                "Error in PME Part interior volume calculation for patno ",
                patno,
                ": "
              ),
              e$message
            )
            NULL
          }
        )
      } else {
        adni_pme_part_volumes <- NULL
      }

      # create empty vectors to store profile area estimates
      patno_adni_area_dim1 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      patno_adni_area_dim2 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      patno_adni_area_dim3 <- vector(
        mode = "numeric",
        length = length(time_values)
      )

      adni_lpme_aug_area_dim1 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      adni_lpme_aug_area_dim2 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      adni_lpme_aug_area_dim3 <- vector(
        mode = "numeric",
        length = length(time_values)
      )

      adni_pme_aug_area_dim1 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      adni_pme_aug_area_dim2 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      adni_pme_aug_area_dim3 <- vector(
        mode = "numeric",
        length = length(time_values)
      )

      adni_pc_part_area_dim1 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      adni_pc_part_area_dim2 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      adni_pc_part_area_dim3 <- vector(
        mode = "numeric",
        length = length(time_values)
      )

      adni_lpme_part_area_dim1 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      adni_lpme_part_area_dim2 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      adni_lpme_part_area_dim3 <- vector(
        mode = "numeric",
        length = length(time_values)
      )

      adni_pme_part_area_dim1 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      adni_pme_part_area_dim2 <- vector(
        mode = "numeric",
        length = length(time_values)
      )
      adni_pme_part_area_dim3 <- vector(
        mode = "numeric",
        length = length(time_values)
      )

      np <- import("numpy")
      o3d <- import("open3d")
      pv <- import("pyvista")

      palette <- colorRampPalette(c(brewer.pal(8, "Spectral")))
      palette_colors <- palette(101)
      time_colors <- palette_colors[ceiling((time_values + 1e-10) * 100)]

      # for each model estimate at each time point, project reconstructed points
      # to coordinate plane and compute area estimate using Delaunay triangulation

      for (time_idx in seq_along(time_values)) {
        x_scale <- patno_adni_centers$max_x[time_idx]
        y_scale <- patno_adni_centers$max_y[time_idx]
        z_scale <- patno_adni_centers$max_z[time_idx]

        temp_data <- patno_adni |>
          filter(time_from_bl == time_values[time_idx]) |>
          select(x, y, z) |>
          as.matrix()

        temp_data_proj_dim1 <- mesh_projection(temp_data, axis = 1)
        temp_data_proj_dim2 <- mesh_projection(temp_data, axis = 2)
        temp_data_proj_dim3 <- mesh_projection(temp_data, axis = 3)

        patno_adni_area_dim1[time_idx] <- temp_data_proj_dim1$area
        patno_adni_area_dim2[time_idx] <- temp_data_proj_dim2$area
        patno_adni_area_dim3[time_idx] <- temp_data_proj_dim3$area

        if (!is.null(adni_lpme_aug_reconstructions)) {
          temp_lpme_reconstructions <- adni_lpme_aug_reconstructions[
            adni_lpme_aug_reconstructions[, 1] == time_values[time_idx],
            2:4
          ]

          temp_lpme_proj_dim1 <- mesh_projection(
            temp_lpme_reconstructions,
            axis = 1
          )
          temp_lpme_proj_dim2 <- mesh_projection(
            temp_lpme_reconstructions,
            axis = 2
          )
          temp_lpme_proj_dim3 <- mesh_projection(
            temp_lpme_reconstructions,
            axis = 3
          )

          adni_lpme_aug_area_dim1[time_idx] <- temp_lpme_proj_dim1$area
          adni_lpme_aug_area_dim2[time_idx] <- temp_lpme_proj_dim2$area
          adni_lpme_aug_area_dim3[time_idx] <- temp_lpme_proj_dim3$area
        }

        if (!is.null(adni_pme_aug_reconstructions)) {
          temp_pme_reconstructions <- adni_pme_aug_reconstructions[
            adni_pme_aug_reconstructions[, 1] == time_values[time_idx],
            2:4
          ]

          temp_pme_proj_dim1 <- mesh_projection(
            temp_pme_reconstructions,
            axis = 1
          )
          temp_pme_proj_dim2 <- mesh_projection(
            temp_pme_reconstructions,
            axis = 2
          )
          temp_pme_proj_dim3 <- mesh_projection(
            temp_pme_reconstructions,
            axis = 3
          )

          adni_pme_aug_area_dim1[time_idx] <- temp_pme_proj_dim1$area
          adni_pme_aug_area_dim2[time_idx] <- temp_pme_proj_dim2$area
          adni_pme_aug_area_dim3[time_idx] <- temp_pme_proj_dim3$area
        }

        if (!is.null(adni_pc_part_reconstructions)) {
          temp_pc_reconstructions <- adni_pc_part_reconstructions[
            adni_pc_part_reconstructions[, 1] == time_values[time_idx],
            2:4
          ]

          temp_pc_proj_dim1 <- mesh_projection(
            temp_pc_reconstructions,
            axis = 1
          )
          temp_pc_proj_dim2 <- mesh_projection(
            temp_pc_reconstructions,
            axis = 2
          )
          temp_pc_proj_dim3 <- mesh_projection(
            temp_pc_reconstructions,
            axis = 3
          )

          adni_pc_part_area_dim1[time_idx] <- temp_pc_proj_dim1$area
          adni_pc_part_area_dim2[time_idx] <- temp_pc_proj_dim2$area
          adni_pc_part_area_dim3[time_idx] <- temp_pc_proj_dim3$area
        }

        if (!is.null(adni_lpme_part_reconstructions)) {
          temp_lpme_part_reconstructions <- adni_lpme_part_reconstructions[
            adni_lpme_part_reconstructions[, 1] == time_values[time_idx],
            2:4
          ]

          temp_lpme_part_proj_dim1 <- mesh_projection(
            temp_lpme_part_reconstructions,
            axis = 1
          )
          temp_lpme_part_proj_dim2 <- mesh_projection(
            temp_lpme_part_reconstructions,
            axis = 2
          )
          temp_lpme_part_proj_dim3 <- mesh_projection(
            temp_lpme_part_reconstructions,
            axis = 3
          )

          adni_lpme_part_area_dim1[time_idx] <- temp_lpme_part_proj_dim1$area
          adni_lpme_part_area_dim2[time_idx] <- temp_lpme_part_proj_dim2$area
          adni_lpme_part_area_dim3[time_idx] <- temp_lpme_part_proj_dim3$area
        }

        if (!is.null(adni_pme_part_reconstructions)) {
          temp_pme_part_reconstructions <- adni_pme_part_reconstructions[
            adni_pme_part_reconstructions[, 1] == time_values[time_idx],
            2:4
          ]

          temp_pme_part_proj_dim1 <- mesh_projection(
            temp_pme_part_reconstructions,
            axis = 1
          )
          temp_pme_part_proj_dim2 <- mesh_projection(
            temp_pme_part_reconstructions,
            axis = 2
          )
          temp_pme_part_proj_dim3 <- mesh_projection(
            temp_pme_part_reconstructions,
            axis = 3
          )

          adni_pme_part_area_dim1[time_idx] <- temp_pme_part_proj_dim1$area
          adni_pme_part_area_dim2[time_idx] <- temp_pme_part_proj_dim2$area
          adni_pme_part_area_dim3[time_idx] <- temp_pme_part_proj_dim3$area
        }
      }

      # create list of estimates for each model estimate
      data_out <- list(
        data = patno_adni,
        area_dim1 = patno_adni_area_dim1,
        area_dim2 = patno_adni_area_dim2,
        area_dim3 = patno_adni_area_dim3
      )

      adni_lpme_aug_out <- list(
        lpme = adni_lpme_aug,
        reconstructions = adni_lpme_aug_reconstructions,
        area_dim1 = adni_lpme_aug_area_dim1,
        area_dim2 = adni_lpme_aug_area_dim2,
        area_dim3 = adni_lpme_aug_area_dim3,
        fit_time = adni_lpme_aug_time
      )
      adni_pme_aug_out <- list(
        pme = adni_pme_aug_list,
        reconstructions = adni_pme_aug_reconstructions,
        area_dim1 = adni_pme_aug_area_dim1,
        area_dim2 = adni_pme_aug_area_dim2,
        area_dim3 = adni_pme_aug_area_dim3,
        fit_time = adni_pme_aug_time_list
      )

      adni_lpme_part_out <- list(
        lpme = adni_lpme_part,
        reconstructions = adni_lpme_part_reconstructions,
        area_dim1 = adni_lpme_part_area_dim1,
        area_dim2 = adni_lpme_part_area_dim2,
        area_dim3 = adni_lpme_part_area_dim3,
        fit_time = adni_lpme_part_times
      )

      adni_pme_part_out <- list(
        pme = adni_pme_part,
        reconstructions = adni_pme_part_reconstructions,
        area_dim1 = adni_pme_part_area_dim1,
        area_dim2 = adni_pme_part_area_dim2,
        area_dim3 = adni_pme_part_area_dim3,
        fit_time = adni_pme_part_times
      )

      adni_pc_part_out <- list(
        pc = adni_pc_part,
        reconstructions = adni_pc_part_reconstructions,
        area_dim1 = adni_pc_part_area_dim1,
        area_dim2 = adni_pc_part_area_dim2,
        area_dim3 = adni_pc_part_area_dim3,
        fit_time = adni_pc_part_times
      )

      adni <- list(
        data = data_out,
        lpme_aug = adni_lpme_aug_out,
        pme_aug = adni_pme_aug_out,
        lpme_part = adni_lpme_part_out,
        pme_part = adni_pme_part_out,
        pc_part = adni_pc_part_out
      )

      # save outputs
      if (!dir.exists(here(paste0("output/adni_area/", patno)))) {
        dir.create(here(paste0("output/adni_area/", patno)), recursive = TRUE)
      }

      print(
        paste0(
          "Saving results for participant ",
          patno,
          ", structure ",
          structure
        )
      )

      saveRDS(
        adni,
        here(paste0(
          "output/adni_area/",
          patno,
          "/",
          structure,
          "_results.RDS"
        ))
      )
      # output TRUE if fitting process completed
      TRUE
    }
  if (verbose == FALSE) {
    stopCluster(cl)
  }
  results <- adni_fit
}
```

With the `preprocess_adni` and `fit_adni` functions defined, the code in the following chunk applies those functions to MRI data from ADNI participants for all structures of interest. This code is included in `code/adni_application.R`, which was in turn run as a batch job on the high performance computing cluster with `adni_lpme.sh`.

```{r}
lhipp_surface <- read_csv(here("data/lhipp_surface_fsl.csv"))
rhipp_surface <- read_csv(here("data/rhipp_surface_fsl.csv"))
lthal_surface <- read_csv(here("data/lthal_surface_fsl.csv"))
rthal_surface <- read_csv(here("data/rthal_surface_fsl.csv"))

adni_info <- read_csv(here("data/adni_info_full.csv")) |>
  distinct()
adni <- read_csv(here("data/adni_info.csv")) |>
  rename(
    image_id = "Image Data ID",
    subid = "Subject",
    date = "Acq Date"
  )

lhipp_processed <- preprocess_adni(lhipp_surface, adni_info)
lhipp_surface <- lhipp_processed$surface
lhipp_centers <- lhipp_processed$centers

rhipp_processed <- preprocess_adni(rhipp_surface, adni_info)
rhipp_surface <- rhipp_processed$surface
rhipp_centers <- rhipp_processed$centers

lthal_processed <- preprocess_adni(lthal_surface, adni_info)
lthal_surface <- lthal_processed$surface
lthal_centers <- lthal_processed$centers

rthal_processed <- preprocess_adni(rthal_surface, adni_info)
rthal_surface <- rthal_processed$surface
rthal_centers <- rthal_processed$centers

write_csv(lhipp_surface, here("data/lhipp_surface_fsl_processed.csv"))
write_csv(rhipp_surface, here("data/rhipp_surface_fsl_processed.csv"))
write_csv(lthal_surface, here("data/lthal_surface_fsl_processed.csv"))
write_csv(rthal_surface, here("data/rthal_surface_fsl_processed.csv"))

lhipp_fit <- fit_adni(
  lhipp_surface,
  lhipp_centers,
  "lhipp",
  verbose = FALSE,
  cores = parallel::detectCores() / 2
)

lthal_fit <- fit_adni(
  lthal_surface,
  lthal_centers,
  "lthal",
  verbose = TRUE,
  cores = parallel::detectCores() / 2
)

rthal_fit <- fit_adni(
  rthal_surface,
  rthal_centers,
  "rthal",
  verbose = FALSE,
  cores = parallel::detectCores() / 2
)
rhipp_fit <- fit_adni(
  rhipp_surface,
  rhipp_centers,
  "rhipp",
  verbose = FALSE,
  cores = parallel::detectCores() / 2
)
```
